RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2015-04-17 manuscripts/manuscript.Rmd
further edits, slimming down/simplifying language in metadata and extending nexml/simmap section

First, when writing a NeXML file the package adds certain basic metadata
automatically if they are absent, using default values consistent with
recommended best practices [@Cranston_2014]. Currently, this includes
naming the software generating the NeXML, a time-stamp of when a tree
was produced, and an open data license.

@@

The `RNeXML` interface described above for built-in metadata annotations
perform this cast automatically behind the scenes, including mapping
metadata attributes to terms in requisite common vocabularies or ontologies. Giving
metadata properties (and where applicable, values) as vocabulary or
ontology terms rather than simple text strings is crucial for allowing
machines to not only parse but also interpret and potentially reason
over their semantics [@Parr2011]. To achieve this benefit for custom metadata
extensions, the user necessarily needs to handle certain technical
details from which the `RNeXML` interface shields her otherwise, in particular
the globally unique identifiers (normally HTTP URIs) of metadata terms
and vocabularies. To be consistent with XML terminology, `RNeXML` calls
vocabulary URIs _namespaces_, and their abbreviations _prefixes_. For
example, the namespace for the Dublin Core Metadata Terms vocabulary
is "http://purl.org/dc/elements/1.1/". Using its common abbreviation
"dc", a metadata property "dc:title" expands to the identifier
"http://purl.org/dc/elements/1.1/title". This URI resolves to a human and
machine-readable (depending on access) definition of precisely what the
term `title` in Dublin Core means. In contrast, just using the text string
"title" could also mean the title of a person, a legal title, the verb
title, etc.  URI identifiers of metadata vocabularies and terms are not
mandated to resolve, but if machines are to derive the maximum benefit
from them, they should resolve to a definition of their semantics in RDF.

@@

Here we illustrate how `RNeXML`'s interface to NeXML's metadata system
can be used to record and share a type of phylogenetic data not taken
into account when NeXML was designed, in this case stochastic character
maps [@Huelsenbeck_2003]. Such data assign certain parts (corresponding
to time) of each branch in a time-calibrated phylogeny to a particular
"state" (typically of a morphological characteristic). The current
de-facto format for sharing stochastic character maps, created by
`simmap` [@Bollback_2006], a widely used tool for creating such maps,
is a non-interoperable modification of the standard Newick tree format.

To express these data as a custom metadata extension, we create metadata
annotations for the `edge` elements in the phylogeny topology. The
metadata need to describe the character state being assigned, and
the duration (in terms of branch length) that the edge spends in
that state. The `NeXML` standard already defines `state` elements for
discrete character traits, which, as most elements in NeXML, also have
unique identifiers by which they can be referenced.

In the following
proposal, we term each series of character state assignments for an
edge a _reconstruction_, and for each state assignment comprising a
reconstruction, we provide the length, the identifier of the character
state, and an ordinal property to record the chronological order of
the assignment. In this manner, a single edge containing one
state change could be annotated as follows.
(the identifier values in the proposal, "cr1", "s1", and
"s2", are identifiers local to a  -- fictitious -- NeXML document,
based on the example simmap tree illustrated in @Revell_2012):

```{r}
 m <- meta("simmap:reconstructions", children = c(
        meta("simmap:reconstruction", children = c(

          meta("simmap:char", "cr1"),
          meta("simmap:stateChange", children = c(
            meta("simmap:order", 1),
            meta("simmap:length", "0.2030"),
            meta("simmap:state", "s2"))),

          meta("simmap:char", "cr1"),
          meta("simmap:stateChange", children = c(
            meta("simmap:order", 2),
            meta("simmap:length", "0.0022"),
            meta("simmap:state", "s1")))
          ))))
```

The example uses a prefix, `simmap`, to group the newly introduced
metadata properties in a vocabulary, for which the `add_namespace()`
method can be used to give a URI as an identifier:

@@

Because the above representation does not modify the NeXML format,
it should be readable -- even if not necessarily interpretable as
a stochastic character mapping -- by any compliant NeXML parser. In
addition, it includes a direct and machine-readable link to metadata
documentation.

Writing out metadata annotations as in the above proposal may seem
tedious. However, by virtue of `RNeXML`'s API it is straightforward to
define functions that encapsulate these details. As a demonstration,
`RNeXML` comes with functions `simmap_to_nexml()` and `nexml_to_simmap()`
that map bidirectionally between NeXML files with `simmap` metadata
according to the above proposal, and the extension of the `ape::phylo`
class devised by @Revell_2012 in the R package `phytools`. The `RNeXML`
simmap vignette illustrates how this allows seamlessly generating and
reading NeXML files containing `simmap` data, and using the various
functions from `phytools` designed for `simmap` objects [@Revell_2012].

>>>>

First, when writing a NeXML file the package adds certain basic metadata
automatically if they are absent, using default values consistent with
recommended best practices [@Cranston_2014]. Currently, this includes
naming the software generating the NeXML, a time-stamp of when a tree
was produced, and an open data license. These are merely default arguments
to `add_basic_meta()` and can be configured.

@@

The `RNeXML` interface described above for built-in metadata allows
users to create precise and semantically rich annotations without 
confronting any of the complexity of namespaces and ontologies.
Nevertheless, advanced users may desire the explicit control over
these semantic tools that takes full advantage of the flexibility
and extensibility of the NeXML specification [@Vos2012; @Parr2011].
In this section we detail how to accomplish these more complex uses
in RNeXML.

Using a vocabulary or ontology terms rather than simple text strings to 
describe data is crucial for allowing
machines to not only parse but also interpret and potentially reason
over their semantics. To achieve this benefit for custom metadata
extensions, the user necessarily needs to handle certain technical
details from which the `RNeXML` interface shields her otherwise: in particular
the globally unique identifiers (normally HTTP URIs) of metadata terms
and vocabularies. To be consistent with XML terminology, `RNeXML` calls
vocabulary URIs _namespaces_, and their abbreviations _prefixes_. For
example, the namespace for the Dublin Core Metadata Terms vocabulary
is "http://purl.org/dc/elements/1.1/". Using its common abbreviation
"dc", a metadata property "dc:title" expands to the identifier
"http://purl.org/dc/elements/1.1/title". This URI resolves to a human and
machine-readable (depending on access) definition of precisely what the
term `title` in Dublin Core means. In contrast, just using the text string
"title" could also mean the title of a person, a legal title, the verb
title, etc.  URI identifiers of metadata vocabularies and terms are not
mandated to resolve, but if machines are to derive the maximum benefit
from them, they should resolve to a definition of their semantics in RDF.

@@

Here we illustrate how `RNeXML`'s interface to NeXML's metadata system
can be used to record and share a type of phylogenetic data not taken
into account when NeXML was designed, in this case stochastic character
maps [@Huelsenbeck_2003]. Such data assign certain parts (corresponding
to time) of each branch in a time-calibrated phylogeny to a particular
"state" (typically of a morphological characteristic). The current
de-facto format for sharing stochastic character maps, created by
`simmap` [@Bollback_2006], a widely used tool for creating such maps,
is a non-interoperable modification of the standard Newick tree format.
This means that computer programs designed to read Newick or NEXUS formats
may fail when trying to read in a phylogeny that includes `simmap` annotations.

NEXML's extensibility avoids this problem.  In NeXML, it is perfectly
possible to add very complex annotations while maintaining a format that
is valid NeXML syntax which can be read by any NeXML parser. To illustrate
how RNeXML facilitates extending the NeXML standard to accommodate such 
complex annotations, we have implemented two functions in the package,
`nexml_to_simmap` and `simmap_to_nexml`.  These functions illustrate
how simmap data can be represented as `meta` annotations on the branch
length elements of a NeXML tree, and provide routines to convert between
this NeXML representation and the extended `ape::phylo` representation
of a `simmap` tree in R that was introduced by @Revell_2012. We encourage
readers to consult the example code in `simmap_to_nexml` to see how
this is implemented.

Extensions to NeXML must also be defined in the file's namespace 
in order to valid.  This provides a way to ensure that a URI providing
documentation of the extension is always included.  Our examples here
use the prefix, `simmap`, to group the newly introduced
metadata properties in a vocabulary, for which the `add_namespace()`
method can be used to give a URI as an identifier:

@@
